Against the Planetary Interpretation of PTFO$\,$8-8695b

### 2020-04-15 paper/ms.tex
update paper

\begin{abstract}
  PTFO$\,$8-8695b, a candidate hot Jupiter in the 7--10$,$Myr old
  Orion-OB1a cluster, might be the youngest hot Jupiter known.
  We inspected data from TESS and Gaia to clarify whether it is
  actually a planet.
  The TESS lightcurve shows that the dominant variability in this
  system is a sinusoidal modulation with a ``long'' period $P_{\rm
  \ell}=11.96\,$hr, likely caused by stellar rotation.  Also present
  is a complex signal, previously identified as the planet candidate,
  that repeats with a ``short'' period $P_{\rm s}= 10.74\,$hr.
  The ``long'' and ``short'' signals beat every 4.48 days.
  Although there is a dip in the short-period signal, ground-based
  photometry from the past decade shows that the orbital phase of the
  dip seems to have instantaneously jumped, at least once, and perhaps
  twice.
  Planets do not ``jump'' in orbital phase.
  Furthermore, the Gaia data show that PTFO$\,$8-8695 is probably
  a photometric binary (the comparison is made relative to the
  kinematic subgroup of 25-Ori for which PTFO$\,$8-8695 is a member).
  Given the available evidence, we believe that PTFO$\,$8-8695 is a
  binary M dwarf system in which one star is showing the ``long''
  rotation signal, and the other star is showing ``transient dipping''
  that has been observed in many young M dwarfs.
  The planetary interpretation seems highly unlikely.
\end{abstract}

@@

Between July 25, 2014 and May 23, 2016, Gaia acquired 121 ``good''
observations of \ptfo, {\it i.e.}, observations that were not strongly
downweighted in the astrometric solution of the source (CITE).
%TODO: give an original explanation of how Gaia works
The Gaia data processing team at XXXINSTITUTION combined the XXX and
YYY into a global astrometric solution, from which the parallax of 
\ptfo\ was derived (CITE).
The star was assigned a Gaia DR2 identifier of 3222255959210123904
(CITE).
In addition, measurements of the stellar proper motion between
observation epochs were made.

@@

\subsection{ Model}

@@

\subsection{ Fitting Results}

@@

As there is no evidence in favor of a blend scenario, we conclude that
both the $P_{\rm s}$ and $P_{\rm \ell}$ signals originate from \ptfo.

\section{Gaia Analysis}
\label{sec:gaia}

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.9\textwidth]{f6.png}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf Positions, kinematics, and HR diagram of \ptfo.}
	Members of the 25$\,$Ori-1 subgroup are shown with black circles, and were identified by \citet{kounkel_apogee2_2018} through clustering on Gaia-DR2 and APOGEE data.
	The ``neighborhood'' (gray circles) is defined as the group of at most $10^4$ randomly selected non-member stars within 5 standard deviations of the mean right ascension, declination, and parallax.
	\label{fig:gaia}
	}
\end{figure*}

We did an analysis.
The result was
Figure~\ref{fig:gaia}.

@@

\subsection{Long period sinusoid}
The standard interpretation of sinusoidal modulations for a
pre-main-sequence M dwarf is that we are observing a rotation period.
There is no evidence to suggest that we are seeing anything different.

>>>>

\begin{abstract}
  PTFO$\,$8-8695b is a candidate hot Jupiter in the 7--10$,$Myr old
  Orion-OB1a cluster. It might be the youngest hot Jupiter known.
  We inspected data from TESS and Gaia to clarify whether it is
  actually a planet.
  The TESS lightcurve shows that the dominant variability in this
  system is a sinusoidal modulation with a ``long'' period $P_{\rm
  \ell}=11.96\,$hr, likely caused by stellar rotation.  Also present
  is a complex signal, previously identified as the planet candidate,
  that repeats with a ``short'' period $P_{\rm s}= 10.74\,$hr.
  The ``long'' and ``short'' signals beat every 4.48 days.
  Although there is a dip in the short-period signal, ground-based
  photometry from the past decade shows that the orbital phase of the
  dip seems to have instantaneously jumped, at least once, and perhaps
  twice.
  Planets do not ``jump'' in orbital phase.
  Furthermore, the Gaia data show that PTFO$\,$8-8695 is probably a
  photometric binary, relative to the subgroup of Orion for which
  PTFO$\,$8-8695 is a member.
  Given the available evidence, we believe that PTFO$\,$8-8695 is a
  binary M dwarf system in which one star is showing the ``long''
  rotation signal, and the other star is showing ``transient dipping''
  that has been observed in many young M dwarfs.
  The planetary interpretation seems highly unlikely.
\end{abstract}

@@

Between July 25, 2014 and May 23, 2016, Gaia acquired 121 ``good''
observations of \ptfo, {\it i.e.}, observations that were not strongly
downweighted in the astrometric solution of the source (CITE).
%TODO: give an original explanation of how Gaia works
The Gaia data processing team at XXXINSTITUTION combined the XXX and
YYY into a global astrometric solution, from which the parallax of 
\ptfo\ was derived (CITE).
The star was assigned a Gaia DR2 identifier of 3222255959210123904
(CITE).
In addition, measurements of the stellar proper motion between
observation epochs were made.
%FIXME sentence is bad

@@

\subsection{Lightcurve Model}

@@

\subsection{Fitting Results}

@@

As there is no evidence in favor of a blend scenario, we conclude that
both the $P_{\rm s}$ and $P_{\rm \ell}$ signals originate from \ptfo,
at least within the resolution of the Gaia-DR2 source catalog.
However, as we shall see, \ptfo itself could still be a binary.

\section{Gaia Analysis \& Photometric Binarity}
\label{sec:gaia}

% \begin{figure*}[t]
% 	\begin{center}
% 		\leavevmode
% 		\includegraphics[width=0.9\textwidth]{f6.png}
% 	\end{center}
% 	\vspace{-0.7cm}
% 	\caption{ {\bf Positions, kinematics, and HR diagram of \ptfo.}
%   Members of the 25$\,$Ori-1 subgroup are shown with black circles,
%   and were identified by \citet{kounkel_apogee2_2018} through
%   clustering on Gaia-DR2 and APOGEE data.  The ``neighborhood'' (gray
%   circles) is defined as the group of at most $10^4$ randomly selected
%   non-member stars within 5 standard deviations of the mean right
%   ascension, declination, and parallax.
% 	\label{fig:gaia}
% 	}
% \end{figure*}
\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.9\textwidth]{f6.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf HR diagram of \ptfo and members of 25$\,$Ori-1.}
  Members of the 25$\,$Ori-1 group (black circles) were identified by
  \citet{kounkel_apogee2_2018} by clustering on six-dimensional
  Gaia-DR2 and APOGEE-2 data.  We defined the reference
  ``neighborhood'' (gray circles) as the group of at most $10^4$
  randomly selected non-member stars within 5 standard deviations of
  the mean 25$\,$Ori-1 right ascension, declination, and parallax.
	\label{fig:gaia}
	}
\end{figure*}

To assess the cluster membership, stellar mass, age, and potential
binarity of \ptfo, we need to identify stars with which it is and is
not coeval.  The simplest way to do this---clustering based on
six-dimensional position and kinematic information---has already been
done by \citet{kounkel_apogee2_2018}.  For simplicity, we considered
the members they identified brighter than $G_{\rm Rp}$ of 16.  This
yielded 149 members of 25$\,$Ori-1.

To define a set of non-associated stars with comparable selection
functions, we defined a reference ``neighborhood'' as the group of at
most $10^4$ randomly selected non-member stars within 5 standard
deviations of the mean 25$\,$Ori-1 right ascension, declination, and
parallax.  We queried these stars using the \texttt{astroquery}
package, which provides a convenient interface to the Gaia archive
(CITE, CITE).  This yielded 1{,}819 neighbor stars.  While some of
these stars may indeed be members of the Orion complex, or even of
25$\,$Ori-1, enforcing this cut on positions and parallaxes ensures
that we are querying stars with comparable amounts of interstellar
reddening.

We examined the resulting five-dimensional right ascension,
declination, proper motions, and parallaxes.  The first point we noted
was that 25$\,$Ori-1 was a clearly defined over-density in each
dimension---the cluster definitely exists, and is different from the
neighborhood.  \ptfo\ was also clearly a member in each of these
projected dimensions.

Given our detection of two separate signals, the issue of whether
\ptfo\ could be a photometric binary was of relatively high concern.
Figure~\ref{fig:gaia} shows the HR diagram from which we can assess
this question.

@@

\subsection{Long period sinusoid}

The standard interpretation of sinusoidal modulations for a
pre-main-sequence M dwarf is that we are observing a rotation period.
There is no evidence to suggest that we are seeing anything different.

This is the dominant signal in the system with 10\% amplitude.
Previous analyses, including the discovery study by
\citet{van_eyken_ptf_2012}, saw the same signal ({\it e.g.}, their
Figure~7).  However it was not previously thought to be astrophysical.
\citet{van_eyken_ptf_2012} identified it as a periodogram peak at
$0.9985 \pm 0.0061\,$days, and ascribed it to the observing cadence,
because of its close correspondence to the sidereal day.  While the
TESS data can show significant reflected light from the Earth
\citep[{\it e.g.},][]{luger_tess_2019}, our pixel-level analysis
showed that the signal is specific to only pixels near \ptfo, and no
other pixels.  We therefore conclude that the signal is astrophysical.
